---
title: "Data Manipulation"
output: html_document
date: "2023-09-23"
---

# Loading Libraries and Setup
Instructs R to print only three lines of each tibble by default.
```{r setup}
library(tidyverse )
options(tibble.print_min = 3)
```

# Loading Data and Cleaning Up Column Names
```{r}
litters_data = read_csv("./data/FAS_litters.csv", 
                        col_types = "ccddiiii")
litters_data = janitor::clean_names(litters_data)

pups_data = read_csv("./data/FAS_pups.csv",
                     col_types = "ciiiii")
pups_data = janitor::clean_names(pups_data)

```

## Select

# Keeping Only Relevant Columns with SELECT 
Can specify the columns I want to keep by naming them in the `select` function:
```{r}
select(litters_data, group, litter_number, gd0_weight, pups_born_alive)
```

Or could specify a range of column to keep:
```{r}
select(litters_data, group:gd_of_birth)
```

Or specify the columns to be removed:
```{r}
select(litters_data, -pups_survive)
```

We can rename variables as part of this process:
```{r}
select(litters_data, GROUP = group, LiTtEr_NuMbEr = litter_number)
```

There are some handy helper functions for `select`; read about all of them using `?select_helpers`. Will likely use `starts_with()`, `ends_with()`, and `contains()` often, especially when variables are named with suffixes or standard patterns.
```{r}
select(litters_data, starts_with("gd"))
```

`everything()` is handy for reorganizing columns without discarding anything:
```{r}
select(litters_data, litter_number, pups_survive, everything())
```

To pull a single variable out of a dataframe, rather than using `select`, use `pull`.

# Renaming Columns with RENAME
If I just want to rename something, can use `rename` instead of `select`. This will rename the variables I care about and keep everything else:
```{r}
rename(litters_data, GROUP = group, LiTtEr_NuMbEr = litter_number)
```

# Rearranging Columns with RELOCATE
Columns specified in the `relocate` function will be moved to the front of the dataframe.
```{r}
relocate(litters_data, litter_number, pups_survive)
```

## Filter

# Include Only Certain Observations with FILTER 
If the data table includes rows I don’t need for the current analysis, I should filter them based on logical expressions using the `filter` function. I often filter using comparison operators `(>, >=, <, <=, ==, and !=)`. Can also use `%in%` to detect if values appear in a set, and `is.na()` to find missing values. The results of comparisons are logical – the statement is TRUE or FALSE depending on the values compared – and can be combined with other comparisons using the logical operators `&` and `|`, or negated using `!`.

Example ways to filter the litters:

* gd_of_birth == 20
* pups_born_alive >= 2
* pups_survive != 4
* !(pups_survive == 4)
* group %in% c("Con7", "Con8")
* group == "Con7" & gd_of_birth == 20

# Omit Missing Observations with DROP_NA
A very common filtering step requires omitting missing observations. Can do this with filter, but prof recommend using drop_na instead.

* `drop_na(litters_data)` will remove any row with a missing value
* `drop_na(litters_data, wt_increase)` will remove rows for which wt_increase is missing.

## Mutate
Can use mutate to change columns or create new ones. 

This example creates a new variable measuring the difference between gd18_weight and gd0_weight and modifies the existing group variable:
```{r}
mutate(litters_data,
  wt_gain = gd18_weight - gd0_weight,
  group = str_to_lower(group)
)
```
A few things in this example are worth noting:

* The new variables can be functions of old variables
* New variables appear at the end of the dataset in the order that they are created
* Can overwrite old variables
* Can create a new variable and immediately refer to (or change) it

## Arrange
Can arrange the rows in your data according to the values in one or more columns:
```{r}
head(arrange(litters_data, group, pups_born_alive), 10)
```

## |>
For example, if I want to clean the column names, remove pups_survive, and create wt_gain. There are a couple of options for this kind of multi-step data manipulation:

* Define intermediate datasets (or overwrite data at each stage)
* Nest function calls

The following is an example of the first option:
```{r}
litters_data_raw = read_csv("./data/FAS_litters.csv",
  col_types = "ccddiiii")
litters_data_clean_names = janitor::clean_names(litters_data_raw)
litters_data_selected_cols = select(litters_data_clean_names, -pups_survive)
litters_data_with_vars = 
  mutate(
    litters_data_selected_cols, 
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group))
litters_data_with_vars_without_missing = 
  drop_na(litters_data_with_vars, wt_gain)
litters_data_with_vars_without_missing
```

Piping allows us to turn this into a sequential chain by passing the result of one function call as an argument to the next function call:
```{r}
litters_data = 
  read_csv("./data/FAS_litters.csv", col_types = "ccddiiii") |> 
  janitor::clean_names() |> 
  select(-pups_survive) |> 
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group)) |> 
  drop_na(wt_gain)

litters_data
```

There are some cases where what you’re piping isn’t going into the first argument. Here, using the placeholder `_` is necessary to indicate where the object being piped should go. 

For example, to regress wt_gain on pups_born_alive, you might use:
```{r}
litters_data |>
  lm(wt_gain ~ pups_born_alive, data = _) |>
  broom::tidy()
```

Will see %>% a lot in code online. It was introduced first, and was very prevalent in tidyverse code for a long time before |> came along.

Command + shift + M is the shortcut for a pipe.