---
title: "Data Manipulation"
output: html_document
date: "2023-09-23"
---

# Loading Libraries and Setup
Instructs R to print only three lines of each tibble by default.
```{r setup}
library(tidyverse )
options(tibble.print_min = 3)
```

# Loading Data and Cleaning Up Column Names
```{r}
litters_data = read_csv("./data/FAS_litters.csv", 
                        col_types = "ccddiiii")
litters_data = janitor::clean_names(litters_data)

pups_data = read_csv("./data/FAS_pups.csv",
                     col_types = "ciiiii")
pups_data = janitor::clean_names(pups_data)

```

# Keeping Only Relevant Columns with SELECT 
Can specify the columns I want to keep by naming them in the `select` function:
```{r}
select(litters_data, group, litter_number, gd0_weight, pups_born_alive)
```

Or could specify a range of column to keep:
```{r}
select(litters_data, group:gd_of_birth)
```

Or specify the columns to be removed:
```{r}
select(litters_data, -pups_survive)
```

We can rename variables as part of this process:
```{r}
select(litters_data, GROUP = group, LiTtEr_NuMbEr = litter_number)
```

There are some handy helper functions for `select`; read about all of them using `?select_helpers`. Will likely use `starts_with()`, `ends_with()`, and `contains()` often, especially when variables are named with suffixes or standard patterns.
```{r}
select(litters_data, starts_with("gd"))
```

`everything()` is handy for reorganizing columns without discarding anything:
```{r}
select(litters_data, litter_number, pups_survive, everything())
```

To pull a single variable out of a dataframe, rather than using `select`, use `pull`.

# Renaming Columns with RENAME
If I just want to rename something, can use `rename` instead of `select`. This will rename the variables I care about and keep everything else:
```{r}
rename(litters_data, GROUP = group, LiTtEr_NuMbEr = litter_number)
```

# Rearranging Columns with RELOCATE
Columns specified in the `relocate` function will be moved to the front of the dataframe.
```{r}
relocate(litters_data, litter_number, pups_survive)
```
















