---
title: "Data Tidying"
output: html_document
date: "2023-09-29"
---

```{r setup}
library(tidyverse)
options(tibble.print_min = 5)
```

Will use the haven package to load the PULSE biomarkers dataset from a .sas7bdat:
```{r}
pulse_df = 
  haven::read_sas("./data/public_pulse_data.sas7bdat") |>
  janitor::clean_names()

pulse_df
```

# `PIVOT_LONGER`

The BDI score is spread across four columns, which correspond to four observation times. We can fix this problem using `pivot_longer`, which essentially "flips" columns with values into new observation rows (but with the same initial ID data).
```{r}
pulse_tidy_data = 
  pivot_longer(
    pulse_df, 
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    values_to = "bdi")

pulse_tidy_data
```

But now, `visit` is an issue. The original column names were informative but we probably don’t need to keep the `bdi_score_ prefix` in each case. Can an additional option in `pivot_longer` to address this:
```{r}
pulse_tidy_data = 
  pivot_longer(
    pulse_df, 
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    names_prefix = "bdi_score_",
    values_to = "bdi")

pulse_tidy_data
```

It's generally bad practice to save intermediate datasets. 

There are also some additional transformations needed to wrap up the data wrangling process:

* Changing bl to 00m for consistency across visits
* Converting visit to a factor variable (or a numeric variable, which could be done with a different call to mutate)

Altogether, then, the code below will import, tidy, and transform the PULSE dataset into a usable format:
```{r}
pulse_df = 
  haven::read_sas("./data/public_pulse_data.sas7bdat") |>
  janitor::clean_names() |>
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    names_prefix = "bdi_score_",
    values_to = "bdi") |>
  mutate(
    visit = replace(visit, visit == "bl", "00m"),
    visit = factor(visit)) 

print(pulse_df, n = 12)
```

## Learning Assessment
In the litters data, the variables gd0_weight and gd18_weight give the weight of the mother mouse on gestational days 0 and 18. Write a data cleaning chain that retains only litter_number and these columns; produces new variables gd and weight; and makes gd a numeric variable taking values 0 and 18 (for the last part, you might want to use recode …).

```{r}

```










