---
title: "Data Import"
output: html_document
---

# Loading Libraries
```{r, echo=FALSE, collapse=TRUE}
library(tidyverse)
library(readxl)
library(haven)
```


# Importing FAS Litters CSV
Let's import the `FAS_litters.csv` csv into a dataframe, then use the janitor package to clean up the column names.

```{r}
litters_df = 
  read_csv("Data/FAS_litters.csv")

litters_df = 
  janitor::clean_names(litters_df)

litters_df

write_csv(litters_df, "Data/litters_cleaned.csv")
```

Now going to import the same dataset using an absolute path.
```{r, include = FALSE}
litters_df_abs = 
  read_csv("~/Documents/Programming/R/Data Science 1/Lectures/Data Wrangling 1/Data/FAS_litters.csv")

litters_df_abs = 
  janitor::clean_names(litters_df_abs)
```

# Importing FAS Pups CSV
Now importing the FAS pups dataset.

```{r}
pups_df = 
  read_csv("Data/FAS_pups.csv")

pups_df = 
  janitor::clean_names(pups_df)

pups_df
```

# Look at Data
```{r}
litters_df

head(litters_df)
tail(litters_df)
```

Can use 'view' to look at the entire dataframe (but will hide from the RMarkdown display becuase it's so big it would make things weird visually)
```{r, eval=FALSE}
view(litters_df)
```

Look at a data summary. The skimr skim function displays a visually pleasant output of the number of missing values per variable, min, max, and a histogram of responses.
```{r}
str(litters_df)

skimr::skim(litters_df)
```

# Options in `read_*`

```{r}
litters_df = 
  read_csv("Data/FAS_litters.csv", skip = 10)
```

Look at NA values.

```{r}
litters_df = 
  read_csv("Data/FAS_litters.csv", na = c("NA",19))
```

Column types
```{r}
litters_df = 
  read_csv("Data/FAS_litters.csv", 
           col_types = 
             cols(
               Group = col_character(),
               `GD of Birth` = col_integer())
           )
```

## Other File Types

Import an xlsx file first.

```{r}
mlb_df = 
  read_excel("Data/mlb11.xlsx")
```

Import a SAS file.
```{r}
pulse_df = 
  read_sas("Data/public_pulse_data.sas7bdat")
```

```{r, eval=FALSE}
view(pulse_df)
```

## Base R

Don't do this

```{r, eval=FALSE}
litters_df = 
  read.csv("Data/FAS_litters.csv")
```




